### 2016-09-29 robust-software.tex
Itemizing feedback

@@

Typically, scientific software is developed and used by a single person,
usually a graduate student or postdoc, and produces excellent results in their
hands~\cite{prins2015}. 
But what happens
when someone else wants to use their software? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
The new user is then faced with two unpalatable options:
hack the existing code to make it work, or start over. Being unable to
replicate results is so common that one publication refers to it as "a rite of
passage"~\cite{baker2016}.

@@

This problem is not unique to bioinformatics, or even to computing~\cite{baker2016}. 
Best practices in software engineering specifically aim to increase software
robustness. However, most bioinformaticians come from related disciplines and
are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
the existing programs and initiatives rarely have the time to cover engineering
topics in-depth, especially since the field is so broad and rapidly
developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
and knowledge are taught, developing robust software is not directly rewarded
in science and funding is difficult to come by~\cite{prins2015}. Some proposed
solutions to this problem, in addition to restructuring educational programs,
include hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
or using specific technical tools like containerization or cloud
computing~\cite{afgan2016,howe2012}. Each of these requires time and, in some
cases, institutional change.

@@

An example of good usage is the Unix \texttt{mkdir} command, which makes
new directories:

@@

Old versions of your software should be available to ensure that results
are reproducible into the far future. A number of mechanisms exist for
controlled release that range from as simple as adding an appropriate
commit message or tag to version control, to official releases alongside
code on Sourceforge, Bitbutcket or Github, to depositing into a
repository like apt, yum, homebrew, CPAN, etc. Choose the method that
best suits the number and expertise of users you anticipate.

@@

Software evolves, but not all software evolves at the same pace.
While more consistent command-line arguments
or more flexible handling of output directories
may make a program better in general,
it can simultaneously make work for someone
who integrated the old version into their own workflow a year or two ago,
and won't see any benefits from upgrading.
Programs' authors should therefore ensure that every version they have released
continues to be available.

@@

In the spirit of code reuse and interoperability, developers often want
to reuse software written by others. 
With a few lines, a call
is made out to the other program, the results are incorporated into the
primary script. Using popular projects reduces the amount of code that
needs to be maintained and adds the strength of vetted software to the
final program.

@@

The way that the second program is invoked can throw up errors.
The program may not be in the user's path, or it may be an
older or newer version and produce results different than expected.
Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
use shell-specific conventions like `*' expansion.
Even Linux-standard functions available vary slightly between
installs. For example, GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

@@

Finally, to ensure support on as many different operating systems as
possible, \textbf{use native functions for starting other processes}, such as
Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
command, and be sure to capture and report the standard error output of
the subprocess to facilitate debugging.

@@

Many bioinformatics applications rely on randomized algorithms to
improve performance or runtimes. As a consequence, results can change
between runs, even when provided with the same data and parameters. By
its nature, this randomness renders strict reproducibility impossible.
Debugging is more difficult. If even the small test set (\#9) produces
different results, new users may not be able to tell whether the software is
working properly. When comparing
results between versions or after changing parameters, even small
differences can confuse or muddy the comparison. And especially when
producing results for publications, grants or diagnoses, any analysis
should be absolutely reproducible.

>>>>

\section*{Notes}

\begin{itemize}

\item
Mention that these rules are necessary but not sufficient for software that needs to interact with services
such as database and web servers.
Such software is often stopped and started under automatic control,
may need to authenticate,
etc.,
all of which are out of scope of this paper.

\item
Mention that reusing software introduces dependencies,
which can bring its own special pain.

\item
``Write documentation'' is not one of our ten rules.
Explain in the conclusion that we think this is really important,
but it's not the software per se.
We also don't think people are unlikely do it during normal development,
and we don't want to recommend practices that people won't actually adopt.

\item
Find somewhere to say, ``Check that all input values are in a reasonable range,
choose reasonable defaults where possible,
and no defaults at all otherwise.''

\item
Explain somewhere that there's an 11th rule for compiled languages:
use standard build tools like Make
rather than home-brewed scripts.

\item
Remind people that a program should print its version number in all of its output,
particularly debugging traces.

\item
Recommend that people \emph{always} install dependencies, even on their own machine,
using their equivalent of \texttt{requirements.txt}
so that they're sure it works.

\item
Point out that even minor changes to code can cause minor changes in output
because of floating-point issues,
which means that getting exactly the same output for the same input and parameters
probably won't work during development
(but should still be a goal for people who've deployed).

\item
Point out that having tests of any kind makes it simpler for outsiders to contribute.

\item
When we tell people not to require root privileges,
we should point out that testing in a virtual machine is a good way to check for this.

\item
References:
  \begin{itemize}
  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-2-15
  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-3-31
  \item https://tspace.library.utoronto.ca/handle/1807/73111
  \end{itemize}

\item
Cite http://hmmer.org/ as an example of scientific software done right.

\item
Suggest that people put all the checking and validation up front,
so that if the program is going to crash it does so right away
(rather than in the middle of the night after 18 hours of runtime).

\item
Should we tell people to create several single-purpose tools
rather than One Script to Rule Them All?

\item
Should we tone down the ``tremor of fear'' remark?

\item
``Keep in mind that if a parameter can be adjusted,
users will want to be able to turn off the feature entirely to have a baseline comparison.''
What's the relationship here between parameter and feature?

\item
Honor local conventions,
e.g.,
use \texttt{TMPDIR} for temporary files.
(The problem is,
where are all of these written down?)

\item
Combine versioning and making previous versions available into a single point
in order to free up space for something else to be rule 10?

\end{itemize}

@@

Typically, scientific software is developed and used by a single person,
usually a graduate student or postdoc, and produces intended results in their
hands~\cite{prins2015}. 
But what happens
when someone else wants to use their software? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
The new user is then faced with two unpalatable options:
hack the existing code to make it work, or start over. Being unable to
replicate results is so common that one publication refers to it as "a rite of
passage"~\cite{baker2016}.

@@

This problem is not unique to bioinformatics, or even to computing~\cite{baker2016}. 
Best practices in software engineering specifically aim to increase software
robustness. However, most bioinformaticians come from related disciplines and
are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
the existing programs and initiatives rarely have the time to cover software engineering
in-depth, especially since the field is so broad and rapidly
developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
and knowledge are taught, developing robust software is not directly rewarded
in science and funding is difficult to come by~\cite{prins2015}. Some proposed
solutions to this problem, in addition to restructuring educational programs,
include hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
or using specific technical tools like containerization or cloud
computing~\cite{afgan2016,howe2012}. Each of these requires time and, in some
cases, institutional change.

@@

An example of good usage is GNU's \texttt{mkdir} command, which makes
new directories:

@@

@@

Software evolves, but not all software evolves at the same pace.
While more consistent command-line arguments
or more flexible handling of output directories
may make a program better in general,
it can simultaneously make work for someone
who integrated the old version into their own workflow a year or two ago,
and won't see any benefits from upgrading.
A program's authors should therefore ensure that every version they have released
continues to be available.

A number of mechanisms exist for
controlled release that range from as simple as adding an appropriate
commit message or tag to version control, to official releases alongside
code on Sourceforge, Bitbucket or GitHub, to depositing into a
repository like apt, yum, homebrew, CPAN, etc. Choose the method that
best suits the number and expertise of users you anticipate.

@@

In the spirit of code reuse and interoperability, developers often want
to reuse software written by others. 
With a few lines, a call
is made out to the other program and the results are incorporated into the
primary script. Using popular projects reduces the amount of code that
needs to be maintained and leverages the work done to check that software.

@@

The way the second program is invoked can throw up errors.
The program may not be in the user's path, or it may be an
older or newer version and produce results different than expected.
Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
use shell-specific conventions like `*' expansion.
Even Linux-standard functions available vary slightly between
installs. For example, GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

@@

Finally, to ensure support on as many different operating systems as
possible, \textbf{use native functions for starting other processes}, such as
Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
command, and be sure to capture and report the output of the subprocess's standard error
to facilitate debugging.

@@

Many applications rely on randomized algorithms to
improve performance or runtimes. As a consequence, results can change
between runs, even when provided with the same data and parameters. By
its nature, this randomness renders strict reproducibility impossible.
Debugging is more difficult. If even the small test set (\#9) produces
different results, new users may not be able to tell whether the software is
working properly. When comparing
results between versions or after changing parameters, even small
differences can confuse or muddy the comparison. And especially when
producing results for publications, grants or diagnoses, any analysis
should be absolutely reproducible.
